A family of accessibility measures

### 2024-08-14 paper/manuscript.qmd
Additions regarding "The Law of Retail Gravitation"

<!--Hansen leans on Stewart's 1948 Population Potential measure. Two crucial things: 1) Stewart set the proportional constant to 1 *for future determination* and 2) Stewart suggests that keeping the exponent value of d_ij as 1 is sufficient. Hansen doesn't address the proportional constant of 1 explicitly in his application and picks an exponent of then 1 for d_ij as it fits empirically better: the issue of units are never addressed. -->
Following this literature, accessibility as the  _"potential of opportunities for interaction"_ defined in @hansen1959 emerged. Today, @hansen1959 is frequently cited as the father of the term's scientific use [@reggianiGuestEditorialNew2011] and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure. @hansen1959 borrowed the "population potential" formulation used in @stewartDemographicGravitationEvidence1948, which closely follows the Newtonian definition of energy potential, and applied it to measure potential of opportunities (for interaction) (@eq-2). ^[The integral of $F_{ij}$ (@eq-1) over distance $d_{ij}$: $U_{ij}=-G M_{i}M_{j} /d_{ij}$. The potential energy of one mass, say $M_i$, is hence $U_i = G M_j / d_{ij}$, when applied to population instead of mass this equation yields 'potential population'. @hansen1959 takes this equation and replaces $M_j$ with _opportunities_ to derive _opportunity potential_ or, as Hansen defines, _potential of opportunities for interaction_.] $S_ij$ is a relative measure of the accessibility at $i$ to an activity located at $j$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size. ^[Notably, @Hansen1959 finds that when the model is empirically calibrated, ${\beta}$ takes a value of between 0.5 to 3, not the value of 1 as consistent in the Newtonian potential energy relationship.]

@@

<!--Accessibility from Hansen to today: very diverse, hard to interpret, and its lack of concern for the proportionality constant.  Our answer: the proportionality constant.-->
Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to the authors knowledge, have not been concerned with explicitly determining a proportionality constant. The majority of studies develop on the breath of opportunity $O_j$ studied (i.e., work, shopping, schools, leisure ()) and their geographic case study selection, the accuracy of parameters that capture distance-decay $f(c_{ij})$ (), and additions to the opportunity $O_j$ attribute such as competitive demand considerations (). However, the difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction, plagues it to this day and contributes to resistance to its application in practice(). The quantity and variations on @eq-3 compounds this issue of interpretability across studies. A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (gravity), utility-, location-, and person- based measures with different components, and @wuUnifyingAccess2020 demonstrates that the concept can be re-unified into its main features: the relationship between travel-cost and corresponding reachable opportunities and demonstrates how previous classifications are special cases of travel-cost or opportunities constraints. From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing - the proportionality constant $G$. The aim of this paper's focus is on the proportionality constant to re-unify accessibility research with its Newtonian gravitational roots.

@@

Derive the various accessibility measures. In my mind:
•	Non-constrained: it does not include a single proportionality constant.
•	Unconstrained/Total constrained: uses a single proportionality constant
•	Singly constrained: spatial availability
•	Composite singly constrained: Balanced floating catchment areas (not truly doubly constrained)
•	Doubly constrained (only when the units of the marginals are the same, therefore of more limited application)

>>>>

<!--Hansen leans on Stewart's 1948 Population Potential measure. Two crucial things: 1) Stewart set the proportional constant to 1 *for future determination* and 2) Stewart suggests that keeping the exponent value of d_ij as 1 is sufficient. Hansen doesn't address the proportional constant of 1 explicitly in his application and picks an exponent other than 1 for d_ij as it fits empirically better: the issue of units are never addressed. -->
Following this literature, accessibility as the  _"potential of opportunities for interaction"_ defined in @hansen1959 emerged. Today, @hansen1959 is frequently cited as the father of the term's scientific use [@reggianiGuestEditorialNew2011] and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure. @hansen1959 borrowed the "population potential" formulation used in @stewartDemographicGravitationEvidence1948, which closely follows the Newtonian definition of energy potential, and applied it to measure potential of opportunities (for interaction) (@eq-2). ^[The integral of $F_{ij}$ (@eq-1) over distance $d_{ij}$: $U_{ij}=-G M_{i}M_{j} /d_{ij}$. The potential energy of one mass, say $M_i$, is hence $U_i = G M_j / d_{ij}$, when applied to population instead of mass this equation yields 'potential population'. @hansen1959 takes this equation and replaces $M_j$ with _opportunities_ to derive _opportunity potential_ or, as Hansen defines, _potential of opportunities for interaction_.] $S_ij$ is a relative measure of the accessibility at $i$ to an activity located at $j$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size. ^[Notably, @Hansen1959 finds that when the model is empirically calibrated, ${\beta}$ takes a value of between 0.5 to 3, not the value of 1 as consistent in the Newtonian potential energy relationship.]

@@

<!--Accessibility from Hansen to today: very diverse, hard to interpret, and its lack of concern for the proportionality constant.  Our answer: the proportionality constant.-->
Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to the authors knowledge, have not been concerned with explicitly determining a proportionality constant. The majority of studies develop on the breath of opportunity $O_j$ studied (i.e., work, shopping, schools, leisure () and their geographic case study selection, the accuracy of parameters that capture distance-decay $f(c_{ij})$ (), and additions to the opportunity $O_j$ attribute such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024]. However, the difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction, plagues it to this day and contributes to resistance to its application in practice(). The quantity and variations on @eq-3 compounds this issue of interpretability across studies. A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (gravity), utility-, location-, and person- based measures with different components, and @wuUnifyingAccess2020 demonstrates that the concept can be re-unified into its main features: the relationship between travel-cost and corresponding reachable opportunities and demonstrates how previous classifications are special cases of travel-cost or opportunities constraints. From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing - the proportionality constant $G$. The aim of this paper's focus is on the proportionality constant to re-unify accessibility research with its Newtonian gravitational roots.

@@

It's all about balancing factors! Lessons from the balancing factors of the gravity model (cool paper that explores this @kirbyNormalizingFactorsGravity1970)

Derive the various accessibility measures. In my mind:
•	Non-constrained: it does not include a single proportionality constant.

•	Unconstrained/Total constrained: uses a single proportionality constant

•	Singly constrained: spatial availability

Origins in retail models, decision model for shops to optimally spatially locate to max. their potential consumer interaction. Starting with reilly's law of gravitational retail (1931) (can't find online), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.

Independently formed in people spatial interaction: "intervening opportunities and competing migrants" @stoufferInterveningOpportunitiesTheory1940 and @stoufferINTERVENINGOPPORTUNITIESCOMPETING1960, "axiomatically"  @weibull_axiomatic_1976, "for multiple-modes, as a rate" @shen1998;for multiple-modes, as the number of opportunities" @soukhovMultimodalSpatialAvailability2024

•	Composite singly constrained: Balanced floating catchment areas (not truly doubly constrained)

•	Doubly constrained (only when the units of the marginals are the same, therefore of more limited application)
ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020
